---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


### Step 1 - Loading and preprocessing the data

Load needed R packages ...
```{r library, echo=TRUE, message=FALSE}
getwd()

library(data.table) 
library(dplyr)
library(ggplot2)
```

Download and unzip the input data file ...
```{r download, echo=TRUE}

if(!file.exists("activity.zip")) {
    temp <- tempfile()
    download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
    unzip(temp)
    unlink(temp)
}
```

Read the unzipped .csv input data file ... 
```{r load_data, echo=TRUE}

df <- read.csv("activity.csv")

str(df) 
```



### Step 2 - What is mean total number of steps taken per day?

Using **Aggregate** fucntion to calculate the total number of steps by day ...
```{r aggregation1, echo=TRUE}

steps_by_day <- aggregate(steps ~ date, df, sum)

head(steps_by_day)
```

Create a **histogram** showing the total number of steps taken each day ...
```{r Plot1, echo=TRUE}

hist(steps_by_day$steps, 
     main   = "Histogram - Number of Steps taken each Day", 
     xlab   = "Number of steps taken each day", 
     ylab   = "Frequency",
     breaks = 20,
     col    = "green")
```

Calculate the **mean** of the total number of steps taken each day ...
```{r mean1, echo=TRUE}

mean1 <- mean(steps_by_day$steps)

mean1
```

Calculate the **median** of the total number of steps taken each day ...
```{r median1, echo=TRUE}

median1 <- median(steps_by_day$steps)

median1
```



### Step 3 - What is the average daily activity pattern?

Calculate the **average** number of steps taken across all days by interval ...
```{r average, echo=TRUE}

avg_steps_by_interval <- aggregate(steps ~ interval, df, mean)

head(avg_steps_by_interval,5)   ## str(avg_steps_by_interval)
```

Let's find the **index** of the 5-minutes interval, on average across all days in the dataset, that contains the maximum number of steps ...
```{r index_max, echo=TRUE}

index_max <- which.max(avg_steps_by_interval$steps)

index_max
```

Then, get the **row** that contains the max values first so we can plot on the chart too ...
```{r row_max, echo=TRUE}

row_max <- avg_steps_by_interval[index_max,]

row_max
```

Now, create a time series plot `type = "l"` of 5-minutes interval (x-axis) and the **average** number of steps taken across all days (y-axis) ...
```{r Plot2, echo=TRUE}

  ggplot(avg_steps_by_interval, aes(x = interval, y = steps)) + 
       geom_line() + 
       geom_point() +
       geom_vline(xintercept = row_max[[1]], 
                  colour = "red", 
                  linetype = "dashed") +
       geom_hline(yintercept = max(avg_steps_by_interval$steps), 
                  colour = "red", 
                  linetype = "dashed") +
       geom_text(aes(row_max[[1]],0, 
                     label = row_max[[1]], 
                     hjust=-0.5), 
                     colour="blue", 
                     size=4) +
       geom_text(aes(0,row_max[[2]], 
                     label = round(row_max[[2]],2), 
                     vjust=-0.5), 
                     colour="blue", 
                     size=4) +  
       ggtitle("Average number of steps by 5-minutes interval \n") +
       labs(x = "\n 5-minutes interval") + 
       labs(y = "Average number of steps \n")
```



### Step 4 - Imputing missing values

The total number of missing values in the dataset is ....
```{r total_missing1, echo=TRUE}

sum(!complete.cases(df))
```

Use the average number of steps per 5-minutes interval ( calculated in previous question ), to replace the **N/A** values in the dataset ... 
```{r, echo=TRUE}

library(magrittr)

df2 <- inner_join(df, avg_steps_by_interval, by="interval") %>% 
                  mutate(steps=ifelse(is.na(steps.x),steps.y,steps.x)) %>%
                  select(date,interval,steps)
```


Let's verify the new dataset ...
```{r total_missing2, echo=TRUE}

sum(!complete.cases(df2))
```

Then, aggregate the number of steps by day for the new dataset ...
```{r aggregation2, echo=TRUE}

steps_by_day2 <- aggregate(steps ~ date, df2, sum)

head(steps_by_day2)
```
  
There is no missing values in the new dataset. So, let's create the **histogram** again ...
```{r Plot3, echo=TRUE}

hist(steps_by_day2$steps, 
     main   = "Histogram - Number of Steps taken each Day", 
     xlab   = "\n Number of steps taken each day \n( without missing values )", 
     ylab   = "Frequency",
     breaks = 20,
     col    = "green")
```

Recalculate the **mean** of the total number of steps taken each day ...
```{r mean2, echo=TRUE}

mean2 <- mean(steps_by_day2$steps)

mean2
```

Recalculate the **median** of the total number of steps taken each day ...
```{r median2, echo=TRUE}

median2 <- median(steps_by_day2$steps)

median2
```

**Note:** The **mean** of the number of steps taken each day remains the same because the missing values is replaced with the average. The **median** is now equal to the **mean**. 



### Step 5 - Are there differences in activity patterns between weekdays and weekends?

Create factor variables **"weekday"** and **"weekday"** in the filled-in missing values dataset  ...
```{r, echo=TRUE}

df3 <- transform(df2, 
                 day_type = ifelse(weekdays(as.Date(steps_by_day2$date)) == "Saturday" | 
                                   weekdays(as.Date(steps_by_day2$date)) == "Sunday", 
                                   "weekend", "weekday")
                 )

str(df3)
```

Make a panel containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis) ...
```{r Plot4, echo=TRUE}

avg_steps_by_interval_day_type <- aggregate(steps ~ interval + day_type, df3, mean)

ggplot(avg_steps_by_interval_day_type, aes(x = interval, y = steps)) + 
      geom_line( col = "blue" ) + 
      facet_wrap(~ day_type, ncol = 1) +
      ggtitle("Average number of steps each day by interval \n( Weekday vs. Weekend ) \n") +
      labs(x = "\n 5-minutes interval") + 
      labs(y = "Average number of steps taken \n")
```
  
  
